---
title: "01-read_filter_normalize_scRNAseq_ALCL"
author: "Putri Ramadani"
date: "2025-05-28"
output: html_document
---

This is documentation for the script `scRNAseq_ALCL_Thymus_RLN_Adult_Ped-merged.Rmd`, which is part of the analysis pipeline for single-cell RNA sequencing data from ALCL patients. 
The script performs data import, filtering, and normalization.
Sample patients:
ALCL001
ALCL002
ALCL003
ALCL004
ALCL005
ALCL006
ALCL007
ALCL008
UHNALCL01
UHNALCL02
UHNALCL03
UHNALCL04

Control samples:
CLC02916 -> Reactive Lymph Node
CLC02917
CLC03483
CLC03510 (removed due to low cell numbers)
CLC04353
EX01176 -> Thymus 
EX01177 (missing in the cluster)
EX01178
EX01179
EX01180

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Load necessary library
```{r}
library(Seurat)
library(pheatmap)
library(ggrepel)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)
```

# Set working directory
```{r}
set.seed(12345)

setwd('~/scRNAseq_ALCL_Thymus_RLN')

lib.info <- readxl::read_xlsx('metadata/lib_ALCL_Thymus_RLN.xlsx')

```


# Data Import & Seurat Object Creation
```{r}
#Create seurat object
sc.obj <- NULL

# Iterate over each row in lib.info to load data
for (i in 1:nrow(lib.info)) {
    i.df <- lib.info[i,]
    i.dir <- paste0(i.df$AnalysisDir)  # Get dataset path
    i.data <- Read10X(data.dir = i.dir)  # Read 10X data
    i.obj <- CreateSeuratObject(counts = i.data, project = i.df$Sample)  # Create Seurat object
    i.obj <- RenameCells(i.obj, add.cell.id = i.df$Sample)  # Rename cells with sample ID
    
    # Merge Seurat objects dynamically
    if (is.null(sc.obj)) {
        sc.obj <- i.obj
    } else {
        sc.obj <- merge(sc.obj, i.obj)
    }
}
```

# Add metadata annotations
```{r} 
lib.info <- lib.info %>% select(Sample, StudyID) # Only keep necessary columns
sc.obj@meta.data <- sc.obj@meta.data %>%
    mutate(
        Sample = orig.ident  # Ensure `orig.ident` matches Sample names
    ) %>%
    left_join(lib.info, by = "Sample")  # Merge metadata based on Sample ID

#Assign Correct Cell Names
rownames(sc.obj@meta.data) <- Cells(sc.obj)
colnames(sc.obj) <- Cells(sc.obj)
```

# QC
```{r}
meta <- sc.obj@meta.data

# add mt%
sc.obj[["percent.mt"]] <- PercentageFeatureSet(sc.obj, pattern = "^MT-")
sc.obj[["percent.rb"]] <- PercentageFeatureSet(sc.obj, pattern = "^RP[LS]")

# These figures will give you a brief overview of the qualities
# check the distribution of nfeature(number of expressed genes), ncount(number of RNA molecues) and mt percent
png(filename = paste0('plots/250625_vlnplot_QC_allsample.png'), width = 1500, height = 600)
print(VlnPlot(sc.obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, group.by = 'StudyID',raster=FALSE))
dev.off()
# warning message: Warning: Default search for "data" layer in "RNA" assay yielded no results; utilizing "counts" layer instead. null device         1
# violin plots show the distribution of nFeature_RNA, nCount_RNA, and percent.mt for each sample in the dataset.
# The x-axis represents the different samples, while the y-axis shows the values for each feature.
# The `nFeature_RNA` represents the number of unique genes detected in each cell,
# `nCount_RNA` represents the total number of RNA molecules detected, 
# and `percent.mt` represents the percentage of mitochondrial gene expression.


#filtering high mt
keep.mito <- sc.obj$percent.mt < 20  # keep cells with mt% < 20%

# in this step, we use 4*MAD to filter outlier cells
keep.nGene <- rep(NA, ncol(sc.obj))
for (s in unique(sc.obj$orig.ident)) {
	s.idx <- which(sc.obj$orig.ident == s)
	s.median <- median(sc.obj$nFeature_RNA[s.idx])
	s.mad <- mad(sc.obj$nFeature_RNA[s.idx])
	s.keep.nGene <- (sc.obj$nFeature_RNA[s.idx] <= s.median + (4 * s.mad)) & (sc.obj$nFeature_RNA[s.idx] >= s.median - (4 * s.mad))
	keep.nGene[s.idx] <- s.keep.nGene
}

# Create some QC plots
qc.df <- sc.obj@meta.data
qc.df <- cbind(qc.df, keep_nGene = keep.nGene, keep_mito = keep.mito, keep = keep.mito & keep.nGene)

png(paste0("plots/250625_barPlot_before_after_qc_allsample.png"), width = 1200, height = 900)
print(
  ggplot(qc.df, aes(x = StudyID, fill = keep)) +
    geom_bar(stat = "count", position = "dodge") +
    scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "red")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 20),
          axis.text.y = element_text(size = 20),
          axis.title.y = element_text(size = 24),
          legend.title = element_text(size = 24),
          legend.text = element_text(size = 20)) +
    labs(x = '', y = 'Number of Cells')
)
dev.off()
# message: null device         1
# meaning that the plot is saved successfully
# The bar plot shows the number of cells before and after QC filtering for each sample.
# The `keep` variable indicates whether a cell passed the QC criteria (TRUE) or was filtered out (FALSE).
# The colors represent the filtering status:
# - Black: Cells that passed QC (keep = TRUE)
# - Red: Cells that were filtered out (keep = FALSE)

# Save filtered seurat object
saveRDS(sc.obj, "data/250625_merged_scobj_qc_preprocessed.rds") 

```

# Data Import & Seurat Object Creation (Remove samples with low cell numbers)
```{r}
# Load the filtered Seurat object
sc.obj <- readRDS("data/250625_merged_scobj_qc_preprocessed.rds")

# Remove samples with low cell numbers
# Define the minimum number of cells required for a sample to be retained
min_cells <- 1000  # Set the threshold for minimum cells

# Create a table of cell counts per sample
cell_counts <- table(sc.obj$orig.ident)

# Identify samples that meet the minimum cell count requirement
samples_to_keep <- names(cell_counts[cell_counts >= min_cells])

# Filter the Seurat object to retain only the samples that meet the criteria
sc.obj <- subset(sc.obj, orig.ident %in% samples_to_keep)

# Save the filtered Seurat object
saveRDS(sc.obj, "data/250625_merged_scobj_qc_preprocessed_filtered.rds")
# Create a bar plot to visualize the number of cells per sample after filtering
png(paste0("plots/250526_barPlot_before_after_qc_allsample_RLN02removed.png"), width = 1200, height = 900)
print(ggplot(qc.df, aes(x = StudyID, fill = keep)) +
	geom_bar(stat = "count", position = "dodge") +
	scale_color_manual(values = c("TRUE" = "black", "FALSE" = "orange")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 20),
        axis.text.y = element_text(size = 20),
        axis.title.y = element_text(size = 24),
        legend.title = element_text(size = 24),
        legend.text = element_text(size = 20))) +
  labs(x = '', y = 'Number of Cells')
dev.off()
# warning message: Warning messages: 1: No shared levels found between `names(values)` of the manual scale and the data's colour values. 2: No shared levels found between `names(values)` of the manual scale and the data's colour values. png  2 

# check the number of cells per sample after filtering
table(sc.obj$orig.ident) # This will show the number of cells per sample after filtering
# output:
#   ALCL001   ALCL002   ALCL003   ALCL004   ALCL005   ALCL006   ALCL007   ALCL008 
#     7699      5915      5846      6825      6475      7076      3219      6013 
#  CLC02916  CLC02917  CLC03483  CLC04353   EX01176   EX01178   EX01179   EX01180 
#     2102      1973      4768      3745      5000      2984      2733      2518 
# UHNALCL01 UHNALCL02 UHNALCL03 UHNALCL04 
#     4932      9421      7407      8505 

# ncol(sc.obj) # This will give you the total number of cells in the filtered Seurat object
# output: [1] 105156

# to know the original number of cells before filtering, you can check the original Seurat object
# sc.obj.original <- readRDS("data/250625_merged_scobj_qc_preprocessed.rds")
# ncol(sc.obj.original) # This will give you the total number of cells in the original Seurat object
# output: 105872

# to check the number of cells of RLN04 sample
# sc.obj.original$orig.ident[sc.obj.original$orig.ident == "CLC03510"]


# How to know how many cells are removed?
# Create a new column in the metadata to indicate whether a sample was kept or removed
qc.df$keep <- ifelse(qc.df$StudyID %in% samples_to_keep
                    , TRUE, FALSE)
# Create a bar plot to visualize the number of cells per sample after filtering min 1000 cells
png(paste0("plots/250526_barPlot_after_qc_allsample_RLN04removed.png"), width = 1200, height = 900)
print(ggplot(qc.df, aes(x = StudyID, fill = keep)) +
    geom_bar(stat = "count", position = "dodge") +
    scale_fill_manual(values = c("TRUE" = "grey", "FALSE" = "orange")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 20),
          axis.text.y = element_text(size = 20),
          axis.title.y = element_text(size = 24),
          legend.title = element_text(size = 24),
          legend.text = element_text(size = 20)) +
    labs(x = '', y = 'Number of Cells')
)
dev.off()

# Interpretation of the bar plot
# The bar plot shows the number of cells per sample after filtering.
# The `keep` variable indicates whether a sample was retained (TRUE) or removed (FALSE).
# The colors represent the filtering status:
# - Grey: Samples that were retained (keep = TRUE)
# - Orange: Samples that were removed (keep = FALSE)
```

# Normalization and clustering
```{r} 
# load the filtered Seurat object
sc.obj_filtered <- readRDS("data/250625_merged_scobj_qc_preprocessed_filtered.rds")

# normalize the data
sc.obj_filtered <- NormalizeData(sc.obj_filtered)

# find the top variable genes
sc.obj_filtered <- FindVariableFeatures(sc.obj_filtered, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
# visualize the top variable genes
top10 <- head(VariableFeatures(sc.obj_filtered), 10)
png(filename = paste0('figures/250625_top10_variable_genes.png'), width = 800, height = 600)
plot <- VariableFeaturePlot(sc.obj_filtered)
plot <- LabelPoints(plot = plot, points = top10, repel = TRUE, size = 5) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    legend.title = element_text(size = 24),
    legend.text = element_text(size = 20)
  ) +
  labs(x = 'Variable Features', y = 'Log2 Variance') +
  ggtitle('Top 10 Variable Genes')

print(plot)
dev.off()

# How to get full ranked variable feature table
# Get metadata of variable features
hvf.info <- HVFInfo(sc.obj_filtered)

# Add gene names as a column
hvf.info$gene <- rownames(hvf.info)

# Order by standardized variance (typically used for ranking)
hvf.ranked <- hvf.info[order(-hvf.info$variance.standardized), ]

# View top 10
head(hvf.ranked, 10)
# Save the ranked variable features to a CSV file
write.csv(hvf.ranked, "data/all_variable_features_ranked.csv", row.names = FALSE)


# scaling the data
sc.obj_filtered <- ScaleData(sc.obj_filtered, verbose = FALSE)

# PCA
sc.obj_filtered <- RunPCA(sc.obj_filtered, npcs = 30, verbose = FALSE)

# UMAP
sc.obj_filtered <- RunUMAP(sc.obj_filtered, reduction = "pca", dims = 1:30, reduction.name = 'umap_GEX')

# SNN
sc.obj_clustered <- FindNeighbors(sc.obj_filtered, reduction = "pca", dims = 1:30)
# clustering. adjust the value of resolution to have a bigger/smaller number of clusters
# a bigger value of resolution will give more clusters, vice versa
sc.obj_clustered <- FindClusters(sc.obj_clustered, reduction = "pca", resolution = 0.8)
# Number of nodes: 105156
# Number of edges: 3752985

# show the clustering results in the UMAP space
png(filename = paste0('figures/250625_umap_Seurat_clusters_before_correction.png'), width = 800, height = 600)
DimPlot(sc.obj_clustered, label = TRUE, label.size = 10, group.by = 'seurat_clusters', reduction = 'umap_GEX', raster=FALSE)
dev.off()

# show the Samples in the UMAP space
png(filename = paste0('figures/250625_umap_StudyID_before_correction.png'), width = 1200, height = 900)
DimPlot(sc.obj_clustered, label = FALSE, label.size = 10, group.by = 'StudyID', reduction = 'umap_GEX', raster=FALSE) +
  theme(legend.text = element_text(size = 30),
        axis.title = element_text(size = 30),
        axis.text = element_text(size = 30) )
dev.off()

# Beacause you read only Sample and StudyID from the metadata, Category1, Category2, or any other useful metadata is permanently dropped at this point.
# Solution:
# You can read the metadata from the original Seurat object and merge it with the filtered Seurat object.
lib.info <- readxl::read_xlsx('metadata/lib_ALCL_Thymus_RLN.xlsx')
colnames(lib.info)
# Add Sample column if not present
sc.obj_clustered$Sample <- sc.obj_clustered$orig.ident

# Merge full metadata from Excel into Seurat
meta <- sc.obj_clustered@meta.data
meta$CellBarcode <- rownames(meta)

# Use dplyr::left_join
meta_merged <- left_join(meta, lib.info, by = "Sample")

# Restore metadata
rownames(meta_merged) <- meta_merged$CellBarcode
meta_merged$CellBarcode <- NULL
sc.obj_clustered@meta.data <- meta_merged
# Now you can access Category1, Category2, or any other metadata

# Plot the UMAP with the updated metadata
# show the Category1 in the UMAP space
png(filename = paste0('figures/250625_umap_Category_before_correction.png'), width = 1200, height = 900)
DimPlot(sc.obj_clustered, label = TRUE, label.size = 10, group.by = 'Category1', reduction = 'umap_GEX', raster=FALSE) +
  theme(legend.text = element_text(size = 30),
        axis.title = element_text(size = 30),
        axis.text = element_text(size = 30) )
dev.off()

# show the Category2 in the UMAP space
png(filename = paste0('figures/250625_umap_Category2_before_correction.png'), width = 1200, height = 900)
DimPlot(sc.obj_clustered, label = FALSE, label.size = 10, group.by = 'Category2', reduction = 'umap_GEX', raster=FALSE) +
  theme(legend.text = element_text(size = 30),
        axis.title = element_text(size = 30),
        axis.text = element_text(size = 30) )
dev.off()

# Save the clustered Seurat object
dir.create("data/Seurat_objects", recursive = TRUE, showWarnings = FALSE)
save(sc.obj_clustered, file = paste0('data/Seurat_objects/sc.obj_clustered.Rdata'))

```

# TO-DO: 
## Do normalization with SoupX
## Do Batch Correction, compare with previous clustering, and save the corrected Seurat object
## Manual cell type annotation